---
title: "Monte Carlo Simulation in R: Predicting the Snooker World Championships"
output: github_document
---

<br>

### Health warning

The code in this repo is intended to demonstrate Monte Carlo simulation when 
applied to a sports tournament - I make no promises of profitability if you
used this for betting. 


### Some background

The World Championships are the biggest event in the snooker calendar. The 
tournament has a pure knockout format. The top 16 players in the world enter the 
first round automatically as seeds while a further 16 join them through a series 
of qualifiers. Unseeded players are randomly drawn to face seeded players with the
paths to the final pre-determined. There is a good diagram on the [tournament Wikipedia page](https://en.wikipedia.org/wiki/2022_World_Snooker_Championship).

Ronnie O'Sullivan won his record-equally 7th championship, though the code 
could be reused for future tournaments.

Each match consists of a number of frames which varies by round:

* the Last 32 is best of 19 frames,
* the Last 16 is best of 25 
* the Quarter Finals are best of 25 
* the Semi Finals are best of 33
* the Final is best of 35

### Data

Two data files are supplied, `players.rds` and `player-ratings.rds`.

The first, `players.rds` contains the players and their match ups for the first
round of the tournament in the order they were drawn in.

```{r echo = FALSE}
library(dplyr)       # for data manipulation
library(readr)       # for reading data
library(here)        # for portable code
library(glue)        # for combining strings
library(matrixStats) # for row max and mins
library(magrittr)    # for piping functions

```

```{r}
players <- "{here::here()}/data/players.rds" %>% 
  glue::glue() %>% 
  readr::read_rds()

print(players)
```

The second dataset, `player-ratings.rds` contains two columns, one with the name
of the player and one with their rating. The rating is roughly their win 
percentage but adjusted for quality of opposition. There are a variety of ways 
you could calculate this rating or approaches you could take which I won't go 
into here but this slide deck [Modelling Outcomes of Professional Snooker Matches](http://www2.stat-athens.aueb.gr/~jbn/conferences/MathSport_presentations/TRACK%20B/B3%20-%20Sports%20Modelling%20and%20Prediction/Collingwood_Snooker.pdf)
is a great place to start.

```{r}
player_ratings <- "{here()}/data/player-ratings.rds" %>% glue() %>% read_rds()

player_ratings %>%
  dplyr::arrange(dplyr::desc(rating)) %>%
  print()
```

### Predicting a frame

Firstly we define a function for predicting the outcome of a frame. The first 
argument `match_rating` is the rating of player 1 minus the rating of player 2.
A positive match rating indicates that player 1 is better while a negative match
rating indicates the opposite. 

The second argument scaling_para scales the probability prediction. I  found a 
good value for this parameter from back testing on frame data to be 0.7. You 
could do this using grid search or R's built in optim or optimise functions.

```{r}
calc_frame_prob <- function (match_rating, scaling_para = 0.7) {
  
  0.5 + scaling_para * match_rating
  
}
```

### Predicting a match

This modelling approach assumes that the probability of a player winning a frame
is the same throughout the match. By making this assumption we can analytically 
calculate match probabilities using the binomial distribution. More information 
about this can be found at [Statistics How To](https://www.statisticshowto.com/probability-and-statistics/binomial-theorem/binomial-distribution-formula/)
.

For each possible score, we will calculate the probability of that score 
occurring, then sum together the probabilities for each possible score to get 
the match probabilities for each player. 

There is a more in depth description of this process in the [Snooker Predictions site](https://snooker-predictions.com/algorithm.html) though the probability of 
winning a frame in this case is found using an ELO model.

We require functions to find possible scores given a best of, calculate the
number of paths to those scores and to bring it all together to calculate the
probability of winning a match.

```{r}
build_possible_scores <- function (best_of) {

  first_to <- ceiling(best_of / 2)
  
  scores <- c(rep(first_to, first_to), 
              0:(first_to - 1), 
              0:(first_to - 1), 
              rep(first_to, first_to))
  
  score_matrix <- matrix(scores, ncol = 2, 
                         dimnames = list(NULL, c("score1", "score2")))

}

calc_num_paths_to_score <- function (possible_scores) {

  x <- matrixStats::rowMaxs(possible_scores) - 1
  y <- matrixStats::rowMins(possible_scores)

  factorial(x + y) / (factorial(x) * factorial(y))

}
```

The following will calculate the probability of winning a match for player 1 
given that players probability of winning a frame and the best of. Matches
played over more frames will favour better players.

```{r}
calc_match_prob <- function (prob_frame, best_of) {

  possible_scores <- build_possible_scores(best_of)
  paths_to_score <- calc_num_paths_to_score(possible_scores)

  prob_winning_scores <- possible_scores %>%
    tibble::as_tibble() %>%
    dplyr::mutate(permutations = paths_to_score,
                  prob_frame,
                  prob_score = permutations * (prob_frame ^ score1) * 
                               ((1 - prob_frame)) ^ score2) %>%
    filter(score1 > score2)
  
  prob_match <- sum(prob_winning_scores$prob_score)
  
}
```

### Simulating the tournament

We now have the ability to predict a match. Now comes the simulation part. 
We will construct a list which contains the structure of the tournament. Each
list element is a data frame / tibble corresponding to a round. From round 2 
onward we have columns player1_prev_match and player2_prev_match which allow us
to allocate winning players to the correct match in the next round.


```{r}
build_tournament <- function (num_players) {

  # Initialise the tournament list
  
  matches_r1 <- tibble(.round = "r1",
                       match_num = 1:(num_players / 2),
                       player1 = NA_character_,
                       player2 = NA_character_)

  matches_r <- matches_r1

  num_rounds <-  log(num_players, base = 2)

  store_rounds <- vector("list", num_rounds)

  store_rounds[[1]] <- matches_r1

  # Loop through each subsequent round
  
  for (i in 2:num_rounds) {

    prev_match_nums <- matches_r$match_num
    prev_round_max_match_num <- max(prev_match_nums)
    prev_round_num_matches <- nrow(matches_r)
    curr_round_num_matches <- prev_round_num_matches / 2

    matches_ri <- tibble(
      .round = paste0("r", i),
      match_num = (1 + prev_round_max_match_num):(prev_round_max_match_num + curr_round_num_matches),
      player1_prev_match = NA_real_,
      player2_prev_match = NA_real_,
      player1 = NA_character_,
      player2 = NA_character_
      )

    # Populate the prev_match columns
    
    prev_match_num_counter <- 0

    for (j in 1:nrow(matches_ri)) {

      for (k in 1:2) {

        prev_match_num_counter <- prev_match_num_counter + 1

        matches_ri[j, 2 + k] <- prev_match_nums[prev_match_num_counter]

      }

    }

    store_rounds[[i]] <- matches_ri
    matches_r <- matches_ri

  }

  return(store_rounds)

}
```

And now we will write a function to simulate a tournament. For each match we 
will use the functions we previous wrote to calculate match probabilities. We 
will generate a random number to decide who won the match and put them into the
next round, recording what stage each player got to. The `results` argument is a
matrix for storing what stage the player got to while `best_ofs` corresponds to
the number of frames played for each round of the tournament.

```{r}
simulate_tournament <- function(tournament, player_ratings, players, results, 
                                best_ofs, scaling_para = 0.7) {
  
  players <- c(tournament[[1]]$player1, tournament[[1]]$player2)
  num_players <- length(players)
  num_rounds <- length(tournament)
  
  for (i in 1:num_rounds) {

    tournament[[i]] <- tournament[[i]] %>%
      dplyr::left_join(player_ratings, c("player1" = "player")) %>%
      dplyr::rename(player1_rating = rating) %>%
      left_join(player_ratings, c("player2" = "player")) %>%
      rename(player2_rating = rating) %>%
      dplyr::mutate(
        match_rating = player1_rating - player2_rating,
        player1_frame_prob = calc_frame_prob(match_rating),
        player1_match_prob = purrr::map_dbl(player1_frame_prob, 
                                            .f = calc_match_prob,
                                            best_of = best_ofs[i]),
        rand = runif(n()),
        winner = if_else(rand < player1_match_prob, player1, player2)
        )
    
    winner_indicator <- rep(0, length(players))
    winner_indicator[players %in% tournament[[i]]$winner] <- 1

    results[,i] <- results[,i] + winner_indicator

    if (i != num_rounds) { # Not required on last round

      tournament[[i + 1]][, "player1"] <- tournament[[i]] %>%
        filter(match_num %in% tournament[[i + 1]]$player1_prev_match) %>%
        dplyr::pull(winner)

      tournament[[i + 1]][, "player2"] <- tournament[[i]] %>%
        filter(match_num %in% tournament[[i + 1]]$player2_prev_match) %>%
        pull(winner)

    }

  }

  return(results)

}

tournament <- build_tournament(32)

player_names <- c(players$player1, players$player2)

results <- matrix(0, nrow = 32, ncol = length(tournament))
rownames(results) <- player_names
colnames(results) <- paste0("r", 1:length(tournament))

print(results[1:5,])
```

We now have all our functions in place and are ready to start simulating. I have
set a seed for the random number generator - you should get the same results for
this simulation. We can see that in this simulation Ronnie O'Sullivan won - there
is a 1 in each column in his row. Further inspection shows Barry Hawkins was 
runner up - he has a 1 in each column of his row except the last one. 

```{r}
best_ofs <- c(19, 25, 25, 33, 35)

tournament[[1]]$player1 <- players$player1
tournament[[1]]$player2 <- players$player2

set.seed(101)

sim1 <- simulate_tournament(tournament, player_ratings, players, results, 
                         best_ofs, scaling_para = 0.7)

print(sim1)
```

That's just one simulation though. The power of this is that you can do it over
and over again, many thousands of times, and the law of large numbers tells us
the results will converge to the true probabilities. I like using the tictoc 
package for timing code and outputting where we are at when a long looping 
procedure is under way. Below I have set `num_sim` to be 10 but you would
want to make this number much bigger, say 100k, which will take you a few hours.


```{r}
num_sims <- 10

results_original <- results

tictoc::tic()

for (i in 1:num_sims) {
  
  results <- results + simulate_tournament(tournament, 
                                           player_ratings, 
                                           players, 
                                           results_original, 
                                           best_ofs, 
                                           scaling_para = 0.7)
  
  
  if (i %% 1000 == 0) { # prints an update every 1000 simulations
    
    tictoc::toc()
    print(i)
    tic()
    
  }
  
  
}

results_prob <- results %>%
  magrittr::divide_by(num_sims) %>%
  as_tibble() %>%
  mutate(player = rownames(results), .before = r1) %>%
  arrange(desc(r5))

results_odds <- results_prob %>% mutate(across(r1:r5, ~ 1 / .))

print(results)
print(results_prob)
print(results_odds)
```

## Improvements

### 1. Speed

There are a host of speed improvements you could make to this. One obvious one
is the calc_match_prob function which we map over for each match. Each time
the function derives each potential score and the number of paths to that score
to sum over. There are multiple matches in most rounds so you technically could
just do this once per round. Doing more things with vectors and matrices is 
likely to be quicker.

### 2. Updating ratings

One thing that I didn't highlight but you probably picked up on is that the 
ratings are static. In reality, a players rating would improve. I can imagine 
this would take some speed out of simulations.

### 3. Ratings

I have't gone into depth about the ratings. There are plenty of sources of data
out there like [Snooker.org](www.snooker.org) or [Cuetracker](https://cuetracker.net). 

### 4. Different stages

You could adapt this to focus on different stages or do simulations part way
through the tournament. 
